---
title: "afternoon_lecture"
author: "Shale"
date: "8/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
```

## Warm Up

```{r}
rent_tidy = us_rent_income %>% pivot_wider(names_from = variable, values_from = estimate) %>%  select(-moe)
```

## Penguins practice

Some important functions to remember

`filter()`
`select()`
`mutate()`
`case_when()`
`group_by() %>% summarize()`


## `dplyr::filter()`
keep or exclude **rows** based on conditions
```{r}
fil_exact = filter(.data = penguins, island == "Torgersen")
# unique() is a good one to remember for testing

fil_combo = filter(.data = penguins, species == "Gentoo" & sex == "female")
# same as: fil_combo = penguins %>% filter(species = "Gentoo", sex == "female")

no_g = filter(.data = penguins, species != "Gentoo")

adelie_185 = penguins %>% filter(species == "Adelie" & flipper_length_mm >= 185)
```

%in% looks for any matches within a vector, returns TRUE if value exists within that row.

```{r}
island_check = penguins %>% filter(island %in% c("Dream", "Torgersen"))
# exclusion with %in% would be !island
```

## `dplyr::select()`
select or exclude **columns**

```{r}
chonk = select(.data = penguins, body_mass_g)

halfsies = penguins %>% select(species, year, bill_length_mm:flipper_length_mm)
# note: this moves the year column to be the second column instead of the last

a_sex = select(.data = penguins, -sex)
# same as a_sex = select(.data = penguins, -7)

fatties = penguins %>% select(species, island, body_mass_g) %>% filter(body_mass_g >= 3600)
```

## `mutate()`
add columns or change existing columns (keeps everything that is already there)

```{r}
# same dataset, convert body mass into kg
penguins_units = penguins_kg %>% mutate(body_mass_kg = body_mass_g / 1000,
                                     flipper_length_cm = flipper_length_mm / 10)

# this doesnt exactly work, probably pattern only accepts one input (not happy with logical | )
# penguins_vowels = penguins %>% mutate(species = str_replace(species, pattern = "a" | "e" | "i" | "o" | "A", replacement = "Y"))


# give mutate() a new column name to create a new column, give it an existing column name to have it overwrite an existing column.
```

## `case_when()`
handy variation on if-else

```{r}
size_cat = penguins %>% mutate(size_category = case_when(flipper_length_mm < 185 ~ "small",
                                                         flipper_length_mm >= 185 ~ "medium"))

bill_cat =  mutate(.data = penguins,
                   bird_bill_bin = case_when(bill_length_mm < 38 ~ "short bill",
                                             bill_length_mm >= 38 & bill_length_mm < 43 ~ "medium bill",
                                             bill_length_mm >= 43 ~ "long bill"))
# last argument to case_when() as a catch-all is TRUE ~ "..."
# note that this will also coerce all NA values if any exist
```

## `group_by() %>% summarize()`
find summary values by groups, returns in summary table

```{r}
 sum_i = penguins %>% group_by(species, island) %>%
  summarise(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE),
            sd.flipper_length = sd(flipper_length_mm, na.rm = TRUE))
```


## practice

```{r}
bionic_penguins = penguins %>%
  filter(body_mass_g >= 3500) %>%
  select(species, sex, flipper_length_mm) %>% 
  group_by(species, sex) %>% 
  filter(!is.na(sex)) %>% 
  summarise(bio_mean = mean(flipper_length_mm, na.rm = TRUE),
            bio_sd = sd(flipper_length_mm, na.rm = TRUE),
            bio_max = max(flipper_length_mm, na.rm = TRUE),
            bio_min = min(flipper_length_mm, na.rm = TRUE))
```

